Four decades of water quality change in the upper San Francisco Estuary

### 2017-01-08 sftrends_manu.tex
achemso style

\documentclass[letterpaper,12pt,oneside]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}
\usepackage[paperwidth=8.5in,paperheight=11in,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{setspace}
\usepackage[colorlinks=true,allcolors=Blue]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{indentfirst}
\usepackage{titlesec}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{outlines}
\usepackage{lineno}
\usepackage{array}
\usepackage{times}
\usepackage{cleveref}
\usepackage{acronym}
\usepackage[position=t]{subfig}
\usepackage{paralist}
\usepackage[noae]{Sweave}
\usepackage{natbib}
\usepackage{array}
\usepackage{pdflscape}
\usepackage{bm}
% \usepackage{showlabels}
\bibpunct{(}{)}{,}{a}{}{,}

% page margins and section title formatting
\linespread{1.5}
\setlength{\footskip}{0.5in}
\titleformat*{\section}{\Large\bf\em}
\titleformat*{\subsection}{\singlespace\large\bf}
\titleformat*{\subsubsection}{\singlespace\normalsize\bf\em}
\titlespacing{\section}{0in}{0in}{0in}
\titlespacing{\subsection}{0in}{0in}{0in}
\titlespacing{\subsubsection}{0in}{0in}{0in}

@@

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\raggedbottom
% \linenumbers
\raggedright
\urlstyle{same}
\setlength{\parindent}{0.5in}
\renewcommand\refname{References \vspace{12pt}}

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$, Phil Trowbridge$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ Version Date:   Thu Dec 15 16:04:34 2016 -0600
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Wastewater treatment plants upstream of and within the Delta are a major source of nutrient loading to the system.  As noted in \citep{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.  Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., \citealt{Cornwell14,Novick15}).  Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}. Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in load contributions before/after upgrade,
\item a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 
\item different nitrogen species will have different changes depending on change in load outputs. 
\end{inparaenum}
The use of \ac{wrtds} to describe downstream effects of \ac{wwtp} upgrades could reveal flow-independent trends that have not been previously described.
%See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}

@@

The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \citep{Cohen98}.  The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \citep{Carlton90,Nichols90,Werner93,Crauder16}.  In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.  Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \citep{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \citep{MacNally10}. The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12b}.

@@

Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in \citet{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change. \cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from \citet{Jabusch16}.  Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).

@@

Emphasize the summer/fall change in the 2000s, why is this?  Check \citep{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN? Relate to conceptual diagram.

@@

Data from \citep{Crauder16}, \citet{Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.  Figure 10 in \citet{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.

Note the decrease in Potamocorbula abundance in 2011, 2012.  These are wet years where abundance/biomass of the clams is driven down by lower salinity.  Contrased wtih the annual chlorophyll trends in the same years, the predicted values are above the flow-normalized trend suggesting an increase in chlorophyll with higher flow.  The potential mechanism is therefore a decrease in clam abundance with high flow that releases phytoplankon from filtration pressure.  This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot). See suggestions in \citet{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.

@@

% supplementary material
\beginsupplement

>>>>

\documentclass[journal = esthag, manuscript = article]{achemso}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}

% \usepackage[version=3]{mhchem}
\usepackage[T1]{fontenc}
\usepackage[colorlinks  =true, allcolors=Blue]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{lineno}
\usepackage{cleveref}
\usepackage{acronym}
\usepackage{paralist}
\usepackage{bm}
\usepackage{fixltx2e}
% \usepackage{showlabels}

@@

\author{Marcus W. Beck}
\affiliation{USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, Gulf Breeze, FL}
\email{beck.marcus@epa.gov}
\phone{+1 (850)9342480}
\fax{+1 (850)9342401}
\author{David Senn}
\author{Thomas Jabusch}
\author{Phil Trowbridge}
\affiliation{San Francisco Estuary Institute, Richmond, CA}

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
% \linenumbers

@@

Wastewater treatment plants upstream of and within the Delta are a major source of nutrient loading to the system.  As noted in \cite{Jassby08}, the Sacramento Regional \ac{wwtp} alone contributes 90\% of the ammonium load to the region.  Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc., \citealt{Cornwell14,Novick15}).  Several \acp{wwtp} have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}. Therefore, a modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in load contributions before/after upgrade,
\item a flow-normalized annual trend at P8 to show a change concurrent with WWTP upgrades, and 
\item different nitrogen species will have different changes depending on change in load outputs. 
\end{inparaenum}
The use of \ac{wrtds} to describe downstream effects of \ac{wwtp} upgrades could reveal flow-independent trends that have not been previously described.
%See \href{http://www.waterboards.ca.gov/centralvalley/board_decisions/adopted_orders/san_joaquin/r5-2008-0086_res.pdf}{here}

@@

The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \cite{Cohen98}.  The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \cite{Carlton90,Nichols90,Werner93,Crauder16}.  In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.  Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \cite{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \cite{MacNally10}. The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}.

@@

Estimated annual percent change of ammonium from 1976-1995 at P8 was more than twice that shown in Ref.~\citenum{Jabusch16}, suggesting that flow effects on the time series lead to an under-estimate of the actual change. \cref{tab:trnds} shows an approximate 5\% increase per year, compared to 2\% in Fig 2 from Ref.~\citenum{Jabusch16}.  Mid-1990s was high flow, which diluted the observed ammonium and removal of this component showed a steady, positive increase in the flow-normalized trend (\cref{fig:p8trnds}).

@@

Emphasize the summer/fall change in the 2000s, why is this?  Check \cite{Cloern07}, showed seasonal changes in early 2000s in chlorophyll (NE Pacific shifted to cool phase), is there a mechanism here with DIN? Relate to conceptual diagram.

@@

Data from \cite{Crauder16,Jassby08} describes phytoplankton community changes in the upper estuary, including chlorophyll response to flow.  Figure 10 in Ref.~\citenum{Jassby08} showed that chlorophyll generally decreased with flow in 1980 but inreased with flow in 2000.

Note the decrease in Potamocorbula abundance in 2011, 2012.  These are wet years where abundance/biomass of the clams is driven down by lower salinity.  Contrased wtih the annual chlorophyll trends in the same years, the predicted values are above the flow-normalized trend suggesting an increase in chlorophyll with higher flow.  The potential mechanism is therefore a decrease in clam abundance with high flow that releases phytoplankon from filtration pressure.  This also explains the positive association of chlorophyll with flow in recent years (bottom right dynaplot). See suggestions in Ref.~\citenum{Alpine92,Parchaso02} regarding flow/grazer relationships in the Bay.

@@

% supplementary material
\begin{suppinfo}
\beginsupplement
